Clustering time series based on probability distributions across temporal granularities

### 2021-11-18 main.Rmd
proofread punctuations

It is common to filter data prior to fitting a supervised classification model using instance selection [@olvera2010review] which removes observations that might impede the model building. For clustering, this is analogous to identifying and removing nuisance observations. Prototype selection is more severe than instance selection, because only a handful of cases is selected. @Cutler1994-hm proposed a method called archetypal analysis which has inspired this approach but the procedure we have used follows @Fan2021-bq. First dimension reduction such as t-SNE, MDS or PCA is used to project the data into a 2D space. Second a few "anchor" customers far apart in 2D space are selected. Additional close neighbors to the anchors are selected. To check the selections relative to the full set of variables, we used a tour linked to a t-SNE layout using the R package `liminal` [@R-liminal]. This ensured that the final sample of clustered customers were also far apart in the high-dimensional space. (See the supplementary materials for further details.)<!--Few of these customers have similar distribution across $moy$ and some are similar in their $hod$ distribution. -->

@@

Clustering of the $24$ prototypes <!--and the data subset consisting of $353$ customers--> was conducted with all three distances, JS-NQT, JS-RS and WPD, and is summarised in Figures \ref{fig:opt-cluster-tsne-jsd}, \ref{fig:groups-4and5} and \ref{fig:summary-plot-wpd}. The t-SNE visualization suggests that there are four well-separated clusters. It is possible that because the representation is only 2D, that the fifth group from the original prototype selection is distinctly different in high dimensions. The $sindex$ plots for the three methods indicate some disagreement: JS-NQT suggests $3$, JS-RS suggests $2$ or $5$ and WPD suggests $3$ or $5$. JS-RS would appear to match the original prototypes with the five cluster solution, but it actually differs. Even though the $sindex$ for JS-NQT suggests three clusters, the five cluster solution more closely matches the original prototypes. The WPD clustering provides a different grouping of the customers, and even though it disagrees with the original prototypes it is a useful grouping.<!--While plotting the clusters, RS-based scaling is applied to each customer, so that no customer dominates the shape of the summarized clusters because of magnitude. The plotting scales are not displayed since we want to emphasize comparable patterns rather than scales. The idea is that a customer in a cluster may have low total energy usage, but their behavior may be quite similar to a customer with high usage with respect to distributional pattern or significance across cyclic granularities.-->

@@

```{r summary-plot-wpd, fig.cap = "Summary plots for three (a) and five (b) clusters from WPD showing the $wpd$ values of each customers across $hod$, $moy$, and $wnwd$ through a parallel coordinate plot. Group P-1 and P-2 correspond to Q-1 and Q-2 respectively. Cluster P-3 splits into Q-3, Q-4 and Q-5. P-1 (customers 16, 18) is characterized by high values of $wpd$ on $wnwd$. P-2 has lower $wpd$ for $hod$ than $moy$ and $wnwd$. P-3 behaves exactly opposite to P-2 with higher $wpd$ for $hod$ compared to $moy$ and $wnwd$. For $5$ cluster solution, this groups gets split into Q3-Q5 characterizing different relative significance over $moy$ and $wnwd$.", fig.height=6}
```

@@

Figure \ref{fig:summary-plot-wpd} shows the $wpd$ values $hod$, $moy$, and $wnwd$ of the $24$ customers, colored by $3$ (a) and $5$ (b) cluster solution as suggested by Figure \ref{fig:opt-cluster-tsne-jsd} through a parallel coordinate plot. The variables (wpd's for different granularities) are 
standardized prior to clustering using WPD. In the display, the variables are sorted according to their separation across groups. This means that $wnwd$ is the most important variable in distinguishing the groups, followed by $hod$ and $moy$ for both (a) and (b).
Group P-1 and P-2 correspond to Q-1 and Q-2 respectively. Cluster P-3 splits into Q-3, Q-4 and Q-5. Customers 16 and 18 are characterized by unusual high values of $wpd$ on $wnwd$ compared to the rest of the customers and hence form a group. This could again be verified from figure \ref{fig:hod-ind-group-png}, where these were the only two customers with a difference in their $wnwd$ behavior. They are represented by P-1. P-2 has lower $wpd$ for $hod$ than $moy$ and $wnwd$. This can also be verified from Table \ref{tab:summary-table-wpd} where $hod$ has very low median values of $wpd$ over the members in the cluster, implying, at least half of these customers do not have an interesting $hod$ pattern. P-3 behaves exactly opposite to P-2 with higher $wpd$ for $hod$ compared to $moy$ and $wnwd$. These are the customers who have some significant pattern across $hod$ and this can again be validated by looking at \ref{fig:hod-ind-group-png}.
For $5$ cluster solution, this groups gets split into Q3-Q5 characterizing different relative significance over $moy$ and $wnwd$. For example, Q-4 and Q-5 have almost no pattern across $moy$, but Q-3 has a $moy$ pattern and thus it is reasonable to split them. The patterns could be different, but they are significant. Q-4 and Q-5 are separated because of their significance across $hod$.

@@

In summary, none of the methods captured the five original prototypes exactly. JS-NQT was almost identical, though, but WPD produced quite a different grouping. This is quite a reasonable result and illustrates both the difficulties of clustering to obtain a particularly expected grouping and the ability to learn unexpected patterns in the data. It is possible that the JS-based distances were distracted by the presence of nuisance variables, levels of the granularities, that do not contribute to clustering. This would also be supported by the results of the validation study, where clustering was less effective in S2 and S3, where only some granularities had differences between levels. Clustering using WPD is expected to produce quite different results because it will group only by overall value of a granularity, not a particular pattern. A cluster summary like Figure \ref{fig:groups-4and5} is not possible because there may be different but equally interesting patterns (e.g. high evening $hod$ and high day time $hod$) in the same cluster. Simply it provides information that across a collection of customers this specific cluster has interesting patterns in a granularity (eg $hod$). One would need to post-process the cluster to separate specific patterns.

>>>>

It is common to filter data prior to fitting a supervised classification model using instance selection [@olvera2010review] which removes observations that might impede the model building. For clustering, this is analogous to identifying and removing nuisance observations. Prototype selection is more severe than instance selection, because only a handful of cases is selected. @Cutler1994-hm proposed a method called archetypal analysis which has inspired this approach but the procedure we have used follows @Fan2021-bq. First, dimension reduction such as t-SNE, MDS or PCA is used to project the data into a 2D space. Second, a few "anchor" customers far apart in 2D space are selected. Additional close neighbors to the anchors are selected. To check the selections relative to the full set of variables, we used a tour linked to a t-SNE layout using the R package `liminal` [@R-liminal]. This ensured that the final sample of clustered customers were also far apart in the high-dimensional space. (See the supplementary materials for further details.)<!--Few of these customers have similar distribution across $moy$ and some are similar in their $hod$ distribution. -->

@@

Clustering of the $24$ prototypes <!--and the data subset consisting of $353$ customers--> was conducted with all three distances, JS-NQT, JS-RS and WPD, and is summarised in Figures \ref{fig:opt-cluster-tsne-jsd}, \ref{fig:groups-4and5} and \ref{fig:summary-plot-wpd}. The t-SNE visualization suggests that there are four well-separated clusters. It is possible that because the representation is only 2D, the fifth group from the original prototype selection is distinctly different in high dimensions. The $sindex$ plots for the three methods indicate some disagreement: JS-NQT suggests $3$, JS-RS suggests $2$ or $5$ and WPD suggests $3$ or $5$. JS-RS would appear to match the original prototypes with the five cluster solution, but it actually differs. Even though the $sindex$ for JS-NQT suggests three clusters, the five cluster solution more closely matches the original prototypes. The WPD clustering provides a different grouping of the customers, and even though it disagrees with the original prototypes it is a useful grouping.<!--While plotting the clusters, RS-based scaling is applied to each customer, so that no customer dominates the shape of the summarized clusters because of magnitude. The plotting scales are not displayed since we want to emphasize comparable patterns rather than scales. The idea is that a customer in a cluster may have low total energy usage, but their behavior may be quite similar to a customer with high usage with respect to distributional pattern or significance across cyclic granularities.-->

@@

```{r summary-plot-wpd, fig.cap = "Summary plots for three (a) and five (b) clusters from WPD showing the $wpd$ values of each customers across $hod$, $moy$, and $wnwd$ through a parallel coordinate plot. Group P-1 and P-2 correspond to Q-1 and Q-2 respectively. Cluster P-3 splits into Q-3, Q-4 and Q-5. P-1 (customers 16, 18) is characterized by high values of $wpd$ on $wnwd$. P-2 has lower $wpd$ for $hod$ than $moy$ and $wnwd$. P-3 behaves exactly opposite to P-2 with higher $wpd$ for $hod$ compared to $moy$ and $wnwd$. For $5$ cluster solution, this groups gets split into Q3-Q5 characterizing different relative significance of $moy$ and $wnwd$.", fig.height=6}
```

@@

Figure \ref{fig:summary-plot-wpd} shows the $wpd$ values $hod$, $moy$, and $wnwd$ of the $24$ customers, colored by $3$ (a) and $5$ (b) cluster solution as suggested by Figure \ref{fig:opt-cluster-tsne-jsd} through a parallel coordinate plot. The variables (wpd's for different granularities) are 
standardized prior to clustering using WPD. In the display, the variables are sorted according to their separation across groups. This means that $wnwd$ is the most important variable in distinguishing the groups, followed by $hod$ and $moy$ for both (a) and (b).
Group P-1 and P-2 correspond to Q-1 and Q-2 respectively. Cluster P-3 splits into Q-3, Q-4 and Q-5. Customers 16 and 18 are characterized by unusual high values of $wpd$ on $wnwd$ compared to the rest of the customers and hence form a group. This could again be verified from Figure \ref{fig:hod-ind-group-png}, where these were the only two customers with a difference in their $wnwd$ behavior. They are represented by P-1. P-2 has lower $wpd$ for $hod$ than $moy$ and $wnwd$. This can also be verified from Table \ref{tab:summary-table-wpd} where $hod$ has very low median values of $wpd$ over the members in the cluster, implying at least half of these customers do not have an interesting $hod$ pattern. P-3 behaves exactly opposite to P-2 with higher $wpd$ for $hod$ compared to $moy$ and $wnwd$. These are the customers who have some significant pattern across $hod$ and this can again be validated by looking at \ref{fig:hod-ind-group-png}.
For $5$ cluster solution, this groups gets split into Q3-Q5 characterizing different relative significance of $moy$ and $wnwd$. For example, Q-4 and Q-5 have almost no pattern across $moy$, but Q-3 has a $moy$ pattern and thus it is reasonable to split them. The patterns could be different, but they are significant. Q-4 and Q-5 are separated because of their significance across $hod$.

@@

In summary, none of the methods captured the five original prototypes exactly. JS-NQT was almost identical, though, but WPD produced quite a different grouping. This is quite a reasonable result and illustrates both the difficulties of clustering to obtain a particularly expected grouping and the ability to learn unexpected patterns in the data. It is possible that the JS-based distances were distracted by the presence of nuisance variables, levels of the granularities that do not contribute to clustering. This would also be supported by the results of the validation study, where clustering was less effective in S2 and S3, where only some granularities had differences between levels. Clustering using WPD is expected to produce quite different results because it will group only by overall value of a granularity, not a particular pattern. A cluster summary like Figure \ref{fig:groups-4and5} is not possible because there may be different but equally interesting patterns (e.g. high evening $hod$ and high daytime $hod$) in the same cluster. Simply it provides information that across a collection of customers this specific cluster has interesting patterns in a granularity (eg $hod$). One would need to post-process the cluster to separate specific patterns.
